### 2014-07-21 Paper_v1.tex
First commit on Publications repository

%\bibliographystyle{plain}
\bibliographystyle{plos2009}

@@

\renewcommand{\rmdefault}{phv}\renewcommand{\sfdefault}{phv}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}

@@

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large Dynamically Dimensioned Particle Swarm Optimization}}
\vspace{0.05in}
{\par \centering \large{Adithya Sagar, Christine Shoemaker and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering \large{School of Chemical and Biomolecular Engineering}}
{\par \centering \large{Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Modeling cell free metabolism}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{BMC~Systems~Biology}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853} 
{\par \centering Email: jdv27@cornell.edu} 
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}

{\noindent \textbf{Keywords:}~Parameter identification, Mathematical modeling}

@@

@@

Coagulation is an archetype biochemical network that is highly interconnected and tightly regulated with multiple positive and negative feedback loops. 
The biochemistry underlying coagulation has been well studied [REFHERE], and reliable experimental coagulation models have been developed [REFHERE].

This makes it an ideal system for mathematical modeling and parameter estimation. Coagulation is regulated by a set of serine proteases also known as coagulation factors and blood platelets. The coagulation factors are generally in an inactive state and are known as zymogens. These zymogens are activated when through certain triggers. These trigger events like injury or trauma or sepsis exposure factors like collagen, tissue factor and von Willebrand factor (vWF) to blood. The exposure of these factors to blood kick-starts a series of convergent cascades that lead to conversion of zymogen prothrombinase to thrombin.   Luan et al. modeled coagulation using coupled non-linear ordinary differential equations with 301 reactions and 192 species.

\section*{Discussion}

>>>>

%\bibliographystyle{plain}
\bibliographystyle{msb}

@@

\renewcommand{\rmdefault}{phv}\renewcommand{\sfdefault}{phv}

@@

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large Modeling and Analysis of Hormone and Mitogenic Signal Integration in Prostate Cancer}}
\vspace{0.05in}
{\par \centering \large{ Katharine V. Rogers, Joseph A. Wayman, Ryan Tasseff, Timothy Chu, Kyriakos Tsahalis, Spencer Davis, Rohit Yallamraju, Gongcheng Lu, Matthew P. DeLisa, 
and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering \large{School of Chemical and Biomolecular Engineering}}
{\par \centering \large{Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Modeling signal transduction in prostate cancer}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Molecular Systems Biology}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853} 
{\par \centering Email: jdv27@cornell.edu} 
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
Prostate cancer is the most common cancer in men and the second leading cause of cancer related death in the United States. 
Androgens, such as testosterone, are required for prostate cancer growth. 
Androgen ablation in combination with radiation or chemotherapy remains the primary non-surgical treatment for androgen-dependent prostate cancer. 
However, androgen ablation typically fails to permanently arrest cancer progression, often resulting in castration resistant prostate cancer (CRPC). 
CRPC is closely related to metastasis and decreased survival. 
In this study, we developed and analyzed a population of mathematical models describing growth factor and hormone signal integration in androgen dependent, intermediate and resistant prostate cancer cells. 
The model describes the integration of two simultaneous extracellular signaling inputs, namely, androgen and growth factors into a G1/S cell cycle checkpoint decision. 
Model parameters were identified from 43 studies in androgen dependent and resistant LNCaP cell lines. 
The model was validated by comparing simulations with an additional 29 data sets from LNCaP cell lines that were not used during training. 
Additionally, data from four drug trials was also used to evaluate the model’s performance.
Sensitivity analysis, conducted over an ensemble of prostate signaling models, suggested that in an androgen free environment general translation and transcription was more sensitive in androgen dependent cells, while in androgen independent cells the PI3K and MAPK pathway species were more sensitive. 
In a constant DHT environment sensitive species were conserved between the cell lines. 
These results suggest targeting the PI3K and MAPK pathways in addition to anti-androgen therapies as a treatment for AIPC.

{\noindent \textbf{Keywords:}~Prostate cancer, signal transduction, mathematical modeling}

@@

%
%Sipuleucel-T is a first generation cancer vaccine \cite{Kantoff2010}, while radium-223 is a targeted alpha emitter for bone metastasis \cite{Parker2013}. 
%Abiraterone is an inhibitor of cytochrome P450 17A1 (CYP17), a critical enzyme in androgen biosynthesis \cite{Sartor2013} while enzalutamide inhibits androgen receptor (AR) translocation to the nucleus, as well as DNA binding and coactivator recruitment \cite{Scher2012}.
%Thus, understanding the operation of the biological pathways which ultimately facilitate the transition to CRPC is vital to the discovery of new therapeutic strategies against this disease.
%The model focuses on the outlaw pathway, in which constitutively or overly-activated receptor tyrosine kinases (RTKs) stimulate AKT and mitogen-activated protein kinases (MAPKs), which in turn activate cytosolic androgen receptor (AR) in the absence of androgen. Interestingly, among the few genes activated AR represses is cellular prostatic acid phosphatase (cPAcP), itself a key regulatory of RTK activation. Thus, the outlaw pathway encodes crosstalk between RTK-dependent kinase and AR activation inside of a positive feedback loop. The model was validated with multiple experiments in LNCaP cell lines. In addition  For example, in response to enzalutamide, an AR inhibitor, the model predicted a decline in PSA expression similar to patient response.
%and there is limited data on combination therapies or the use of biomarkers to develop personalized treatment strategies.

Prostate cancer (PCa) is the most commonly diagnosed cancer and the second leading cause of cancer-related death in men in the United States \cite{Siegel2013}. 
Initially, PCa cells depend upon the activation of cytosolic androgen receptors (AR) by androgen hormones, such as testosterone, for survival and growth. 
Androgen ablation in combination with radiation or chemotherapy remains the primary non-surgical treatment for androgen-dependent prostate cancer (ADPC) \cite{Huggins1967}. 
However, androgen ablation typically fails to permanently arrest cancer progression as malfunctioning cells eventually lose androgen sensitivity and proliferate without hormone. 
The loss of androgen sensitivity results in castration resistant prostate cancer (CRPC), a phenotype closely linked with metastasis and greatly reduced survival \cite{Harris2009}.
Currently, there are six approved treatments that demonstrate a survival advantage in patients with metastatic CRPC, each of these target diverse aspects of the disease \cite{Sartor2013}.
The taxane family members docetaxel and cabazitaxel interact with microtubule stability \cite{Tannock2004,Bono2010}, while abiraterone \cite{Sartor2013} or enzalutamide \cite{Scher2012} interfere with androgen signaling by blocking androgen formation or nuclear translocation, respectively.
Other approved treatments are non-specific to PCa. 
For example, general treatments such as sipuleucel-T, a first generation cancer vaccine \cite{Kantoff2010}, 
and radium-223, an alpha emitter which targets bone metastasis \cite{Parker2013}, are both approved to treat CRPC.
Unfortunately, regardless of the therapeutic approach, the survival advantage of these treatments is typically only a few months. 
Thus, understanding the molecular basis of the loss of androgen sensitivity in CRPC could be an important step for the development of the next generation of therapies with a prolonged survival
advantage.

Androgen-induced proliferation and survival depends upon many coordinated signal transduction and gene expression events. 
Androgen Receptor (AR) is part of the nuclear hormone receptor superfamily, which includes other important cancer targets such as progesterone receptor (PR) and estrogen receptor (ER) in breast cancer \cite{Aranda2001}. 
Nuclear hormone receptors act as ligand dependent transcription factors interacting with specific DNA sequences of target genes as either monomers, heterodimers, or homodimers; AR, PR, and ER act as homodimers. 
In the case of AR these specific DNA sequences are known as androgen response elements (ARE) \cite{Mangelsdorf1995}.
In the absence of androgen, AR is predominately found in the cytoplasm bound to heat shock proteins (HSP) \cite{Prescott2006}.
%Nuclear Hormone Receptors (Reviewed \cite{Aranda2001}) and AR in disease reviewed in \cite{Matsumoto2013}
Androgen, either testosterone or testosterone metabolites such as 5$\alpha$-dihydrotestosterone (DHT), enter prostate cells and interact with the cytosolic androgen receptor (AR). 
The interaction of DHT with AR promotes the dissociation of AR from chaperones such as HSP \citep{Pratt:1997kx} and its subsequent dimerization, phosphorylation and translocation to the nucleus (reviewed by Brinkmann \textit{et al.} \citep{Brinkmann1999}). 
Activated nuclear AR drives a gene expression program broadly referred to as androgen action, that promotes both proliferation and survival.
In addition to many genes including itself, activated nuclear AR promotes the expression and secretion of prostate specific antigen (PSA), arguably the best known PCa biomarker \citep{Feldman2001}.
PSA is commonly used as a prostate cancer indicator, although its prognostic ability is controversial \citep{Attard2009,Hoffman2011,Moyer2012}. 
In CRPC, AR signals in the absence of androgens. 
Androgen dependent (AD) prostate cells can become castration resistant (CR) through several possible mechanisms, including constitutively amplified AR expression and altered AR sensitivity to testosterone or other non-androgenic molecules \citep{Feldman2001}. 
In this study, we focused on the aberrant activation of AR by kinase signaling cascades, sometimes called the outlaw pathway. 
Outlaw pathway activation is driven by over-activated receptor tyrosine kinases (RTKs), a common pathology in many cancer types including PCa \citep{Slamon1989,Craft1999}. 
RTKs stimulate downstream kinases, including the AKT and mitogen-activated protein kinase (MAPK) pathways, which promote AR phosphorylation and dimerization in the absence of an androgen signal \citep{Craft1999,Yeh1999}. 
Interestingly, among the few genes activated AR represses is cellular prostatic acid phosphatase (cPAcP), itself a key regulatory of RTK activation \cite{Veeramani2005}. 
Thus, in CRPC the androgen program is initiated without the corresponding extracellular hormone cue, potentially from crosstalk between growth factor and hormone receptor pathways.

In this study, we developed a mathematical model of growth factor and hormone signal integration in androgen dependent, intermediate and resistant prostate cancer cells. 
We used this model to better understand which components and processes were differentially important in AD versus CR cells.
The new model architecture was a significant advance over our previous prostate signaling model \citep{Tasseff2010}. 
We added the regulated expression of ten additional proteins, including the cell cycle restriction point proteins cyclin D (and the differentially spliced variants cyclin D1a and cyclin D1b), cyclin E, cyclin-dependent kinase inhibitor 1A (p21Cip1), and cyclin-dependent kinase inhibitor 1B (p27Kip1). 
Also, we included the Rb/E2F pathway, expanded our description of the activation of the mammalian target of rapamycin (mTOR) protein and its role in translation initiation, and included the regulation of AR action by cyclin D1a and E2F. 
However, this upgraded architecture, while increasing the biological scope of the model, also expanded the number of unknown model parameters. 
To estimate these parameters, we used multiobjective optimization in combination with dynamic and steady-state data sets generated in AD, intermediate and CR LNCaP cell lines. 
We identified a population of approximately N = 5000 models (from well over a million candidate models) which described both AD and CR data sets using a single model structure.
Furthermore, we tested the model using an additional 29 LNCaP data sets not used for model training, along with four drug studies.
We analyzed the model population using sensitivity and robustness analysis to uncover differentially important mechanisms in AD versus CR cell lines. 
In the presence of androgen, the sensitivity profile was similar between AD and CR cells. 
Components of the MAPK and PI3K pathways were highly fragile, irrespective of the level of androgen dependence. 
However, in the absence of androgen, there were 609 statistically significant shifts in species sensitivity between AD and CR cells.
Of these, 108 were larger than one standard deviation above the mean.
In CR cells, HER2 activation of the MAPK and PI3K pathways was significantly more important, as was AR activation through the MAPK pathway. 
On the other hand, components of the translation and transcription infrastructure were differentially more important in AD cells in the absence of androgen.  
Taken together, our analysis suggested that independently targeting the PI3K or MAPK pathways in combination with anti-androgen therapies could perhaps be an effective treatment strategy for CRPC.

%(323 more and 287 less sensitive)
%In this study, we rank-ordered the importance of the molecular components controlling the transition from ADPC to CRPC.
%Toward this, we developed and analyzed an ensemble of mechanistic mathematical models of the signal transduction processes governing the 
%transition from androgen dependent to castration resistant prostate cancer.
%The objective of this study was to rank-order the importance of components in CRPC and ADPC using computational tools. 
%We formulated and analyzed an ensemble of mechanistic mathematical models of the signal transduction processes in CRPC and ADPC 
%that was an extension of our previous work .
%In particular, we focused on the outlaw pathway and the possible crosstalk between RTK-dependent kinases and AR activation. 
%First, we expanded the original model by adding ten  
%Third, we added additional co-activators, such as cyclin E, and inhibitors of AR, such as
%ODE models typically contain a large number of unknown model parameters, which must be estimated  
%Due to uncertainty in parameter estimation and to account for differences between individual cells and cell sublines, an ensemble of parameter sets was used to model the system. 
%Proteins that were not modeled at a gene regulatory level were generated using zeroth order reactions. 
%All protein and RNA species were degraded with first order reaction rates. 
%To obtain an ensemble of parameter sets, we used multiobjective optimization. 
%Relative concentration values were quantified 
%Training data came from experiments on LNCaP sublines. 
%We used , previously developed in the lab \cite{Song2010}, to determine a diverse ensemble of models. 
%C-33 cells, having undergone fewer than 33 passages, are considered AD while C-51 cells, having between 34 and 80 passages, are considered an intermediate prostate cancer cell line. 
%C-81 (passage numbers beyond 80) display a CR phenotype \cite{Lin2000}. 
%http://endo.endojournals.org/content/140/11/5054.long or http://endo.endojournals.org/content/140/11/5054.long#ref-19

@@

\subsection*{Estimating an ensemble of prostate signaling models.} 
We modeled the integration of growth factor, cell cycle and hormone signaling pathways in AD and CR LNCaP cells (Fig. \ref{fg:NetworkSchematic}). 
The signaling architecture was hand curated from over 80 primary literature sources in combination with biological databases.
The model equations were formulated as a system of ordinary differential equations (ODEs), where biochemical reaction rates were modeled using mass action kinetics.
ODEs and mass action kinetics are common modeling tools \citep{Wayman2013}, however, ODEs have the disadvantage of requiring estimates for unknown model parameters.
Many techniques have been developed to estimate ODE model parameters, often from noisy and sparse experimental data \citep{Moles:2003dn}.
Typically these identification problems are underdetermined, hence no unique parameter values can be estimated \citep{Villaverde:2014kl}.
Thus, instead of estimating a single yet highly uncertain parameter set, we estimated an ensemble of possible parameter sets using the Pareto Optimal Ensemble Techniques (POETs) algorithm \citep{Song2010}.
POETs uses a combination of simulated annealing and local optimization techniques coupled with Pareto optimality-based ranking to simultaneously optimize multiple objective functions.
Starting from an initial best fit set, we estimated the 1687 unknown model parameters (1674 kinetic parameters and 13 non-zero initial conditions) using 43 \textit{in vitro} data sets taken from AD, intermediate and CR LNCaP cells (Table~\ref{objective_table}). 
Each of the 43 training data sets was a separate objective in the multiobjective calculation. 
The training data were steady-state or dynamic immunoblots from which we extracted relative species abundance from their optical density profiles. 
POETs sampled well over a million possible parameter sets, from which we selected N = 5000 sets for further analysis. 
Over the ensemble, the coefficient of variation (CV) of the kinetic parameters spanned 0.5 - 5.8, with 33\% of the parameters having a CV of less than one. 
As a control, we also performed simulations for R = 100 random parameter sets to compare against the ensemble generated by POETs.

The ensemble of PCa models recapitulated training data in both AD and CR cell lines with only two experimentally justified parameter changes (Fig. \ref{fg:ColorPlot} and Fig. \ref{fg:Training}). 
Data from the LNCaP clones C-33 (dependent), C-51 (intermediate), and C-81 (resistant) \cite{Lin2000,Igawa2002,Horoszewicz1983} along with the CR LNCaP cell lines LNCaP-Rf \cite{Murillo2001}, LNCaP-AI \cite{Chen2011} and LNAI \cite{Graff2009} were used for model identification. 
To simulate the effective difference between LNCaP cell lines, the parameter controlling the maximum rate of PAcP gene expression was scaled by 0.1 and 0.5, respectively, for the C-81 and C-51 cell-lines compared to C-33. 
This modification was based upon steady-state PAcP data from the three LNCaP clones \cite{Lee2003}. 
Similarly, the expression of p16INK4 was adjusted in accordance with the study of Lu \textit{et al.} \citep{Lu1997}. 
These two parameters were the only adjustable parameter differences between AD and CR cells. 
To simulate an increased mTOR activation in the presence of a DHT stimulus, we added a first order activation term for mTOR activation with a DHT stimulus. Androgens have been shown to increase expression of proteins involved in cellular metabolism, which may lead to an increase in mTOR activation \cite{Xu2006}.
The model fit 36 of the 43 training objectives for greater than 40\% of the ensemble members (Fig. \ref{fg:ColorPlot}A). 
Conversely, only 10 of the 43 training objectives were captured with the random parameter control (Fig. \ref{fg:ColorPlot}B). 
The model captured the crosstalk between RTK activation and androgen action (Fig. \ref{fg:Training}). 
The model described DHT-induced PSA expression in both C-33 (Fig. \ref{fg:Training}A) and C-81 (Fig. \ref{fg:Training}B) cells. 
Interestingly, simulations with the HER2 inhibitor AG879 recapitulated decreased PSA expression in C-81 cells (Fig. \ref{fg:Training}C) in the absence of androgen stimulation.
AR action decreased the PAcP mRNA message (Fig. \ref{fg:Training}D), presumably leading to increased HER2 activity. 
The model also recapitulated the integration of androgen action with AR expression, G1/S cell cycle protein expression and AKT phosphorylation. 
For example, the model captured AR-induced AR expression following a DHT stimulus (Fig. \ref{fg:Training}H).
Conversely, the transcription factor E2F inhibits AR transcription in LNCaP cells (Fig. \ref{fg:Training}I). 
Other cell cycle proteins were also integrated with androgen action.
For example, the cyclin D1 abundance increased in CR compared to AD cells in the absence of androgen (Fig. \ref{fg:Training}E), while DHT induced p21Cip1 expression in C-33 cells (Fig. \ref{fg:Training}F).
The level of phosphorylated AKT was also increased in higher passage number cells (Fig. \ref{fg:Training}G).

%We used a single model architecture to describe the behavior of both AD and CR LNCaP cells, 
%Two new additions to the model were the cell cycle protein p21Cip1 and transcriptional regulation of AR. 
%We were able to capture training data for both mechanisms. 
%AR transcriptionally upregulates its own expression \cite{Lin2003}\cite{Grad1999}. 
%We captured the increase in AR concentration due to AR activation, and thus increased transcriptional activity, upon  
%Higher basal AKT phosphorylation leads to increased activation of the PI3K/AKT pathway in PCa cells with less androgen dependence \cite{Lin2003}. 
%\cite{Davis2006}

%An additional protein we considered in validation that was present in training was phosphorylated SHC-transforming protein 1 (Shc).
%Simulations predicted a similar increase in Shc phosphorylation after DHT stimulus (Fig. \ref{fg:Validation}D) \cite{Veeramani2005Onc}. 
%The Prostate Cancer Working Group (PCWG) updated outcome measurements for castration resistant prostate cancer studies in 2008 \cite{Scher2008}. 
%The primary endpoint in prostate cancer studies is overall survival, while PSA response is often a secondary endpoint. 
%Figure \ref{fg:Validation}I shows fold change of PSA of ensemble members due to the application of sorafenib.

%(Fig. \ref{fg:Validation}H)

\subsection*{Validation revealed missing network structure.} 
The model was validated against 29 \textit{in vitro} and four \textit{in vivo} studies (Table~\ref{prediction_table}). 
For 15 of the 29 cases, greater than 40\% of the ensemble was qualitatively consistent with the experimental data (Fig. \ref{fg:ColorPlot}C). 
However, for the random parameter control, only 7 of the 29 cases were satisfied (Fig. \ref{fg:ColorPlot}D). 
We correctly predicted positive feedback between HER2 auto-activation and androgen action (Fig. \ref{fg:Validation}A and Fig. \ref{fg:Validation}B). 
We also captured the dose-dependence of AR abundance on DHT (Fig. \ref{fg:Validation}C). 
In addition to the cell line studies, we simulated the outcome of enzalutamide, lapatinib, and sorafenib clinical trials in AD and CRPC patients. 
The trial end points were the reduction in PSA expression relative to an untreated baseline.  
Enzalutamide acts on AR by inhibiting its nuclear translocation, DNA binding, and coactivator recruitment \cite{Scher2012}. 
In the enzalutamide trial, 54\% of the patients that received the drug showed a PSA decline of $\geq$ 50\% while 25\% showed a decline $\geq$ 90\%. 
We simulated enzalutamide exposure by reducing the rate constants governing activated AR binding to nuclear importer, cyclin E, and CDK6 to 1\% of their initial values. 
Consistent with the trial, 62\% of ensemble members showed a $\geq$ 50\% decline in PSA abundance, while 14\% showed a $\geq$ 90\% decline (Fig. \ref{fg:Validation}G). 
The second trial we simulated involved exposure of CRPC patients to sorafenib. 
Sorafenib is a kinase inhibitor with activity against Raf, vascular endothelial growth factor receptor (VEGFR), platelet-derived growth factor receptor (PDGFR), c-kit and c-Ret \cite{Dahut2008}. 
We considered only the effects of sorafenib on the protein kinase Raf, as VEGFR, PDGFR, c-kit and c-Ret were not included in the model. 
None of the 22 patients in the sorafenib study showed a PSA decline of $>$ 50\%. 
However, our simulations showed that approximately 55\% of the ensemble members had a PSA decline of $\geq$ 50\%. 
The last drug we considered was lapatinib, an inhibitor of epidermal growth factor receptor (EGFR) and HER2 tyrosine kinase activity \cite{Liu2013}. 
Two lapatinib drug trials were considered: one in which patients had CRPC and one in which patients had biochemically relapsed ADPC \cite{Whang2013,Liu2013}. 
In the CRPC lapatinib drug trial, two of the 21 enrolled patients had a PSA response $\geq$ 47\%  \cite{Whang2013}. 
For the CRPC case, our model showed 26.5\% of ensemble members with a PSA response $\geq$ 47\%. 
Of the 35 patients enrolled in the ADPC lapatinib study, no PSA decreases was observed \cite{Liu2013}. 
In this case, our model showed 9.2\% of ensemble members with a PSA response $\geq$ 50\%.
Although no response to lapatinib was seen in ADPC clinical trials, \textit{in vitro} AD LNCaP experiments showed decreased PSA expression in response to lapatinib, most notably with the addition of DHT \cite{Liu2005}.

Validation and training failures suggested the original signaling architecture was missing critical components. 
Several of the failed training and validation simulations involved the response of the network to epidermal growth factor (EGF) stimulation. 
For example, Chen \textit{et al.} showed that HER2 phosphorylation increased within five minutes following EGF stimulation of LNCaP-AI cells \citep{Chen2011}. 
We predicted no connection between HER2 phosphorylation and EGF stimulation on this short timescale (Fig. \ref{fg:Validation}E). 
Interestingly, we initially neglected the heterodimerization of HER2 with other ErbB family members in order to simplify the model.
However, Chen \textit{et al.} suggested that HER2-EGFR heterodimerization could be an important factor in EGF-driven activation of HER2 \cite{Chen2011}. 
We tested this hypothesis by developing a new model that included HER2 and EGFR heterodimerization. 
We set the rate constants governing the assembly of HER2/EGFR heterodimers equal to EGFR homodimer assembly; all other parameters were unchanged.
This was a reasonable first approximation, as the affinity of HER2/EGFR heterodimerization and EGRF homodimerization is thought to be similar \cite{Hendriks2003}. 
With the inclusion of HER2-EGFR heterodimerization, we qualitatively fit the EGF-induced HER2 activation case and improved our training for experiments that involved an EGF stimulus, e.g., cyclin D mRNA and protein abundance following an EGF stimulus in C-33 cells (Fig \ref{fg:ColorPlot}A and C, white pixels).

%in diverse environments (i.e., DHT, no DHT, drug) 
%Stimuli included the presence or absence of DHT, presence of therapeutics, and also EGF. 
%PIP3 recruits AKT and PDK1 to the membrane where PDK1 activates AKT \cite{Vivanco2002}. 
%Thus, the PIP3 and AKT interaction is required in the model for AKT phosphorylation to occur.

\subsection*{Sensitivity analysis identified differentially important features of the prostate architecture.} 
We used sensitivity analysis to identify important signaling components in AD versus CR cells (Fig. \ref{fg:Sensitivity}). 
We calculated first order steady-state sensitivity coefficients under different stimuli for 500 parameter sets selected from the ensemble. 
Signaling components were rank-ordered based upon analysis of their sensitivity coefficient values. 
In the presence of DHT, the sensitivity profile was similar for AD versus CR cells, with only a few differences (Fig. \ref{fg:Sensitivity}B). 
The top 2\% of sensitive species, regardless of androgen dependence, involved components from the MAPK and PI3K pathways. 
In particular, activated Ras, Raf, phosphorylated MEK, as well as PIP3 localized AKT, phosphorylated AKT, and PI3K were sensitive in both AD and CR cells. 
Species involving PAcP and p16INK4 were more sensitive in AD cells, which was expected since the expression of these two proteins were the only parameters changed between AD and CR cells. 
Other species such as E2F, cyclin E, and DHT-activated AR were also more sensitive in AD cells. 
On the other hand, HER2-Grb2-Gab activation of PI3K and AKT inhibition of RAF were more sensitive in CR cells.

The importance of signaling components varied with androgen dependence in the absence of DHT (Fig. \ref{fg:Sensitivity}A). 
There were 609 statistically significant shifts in species sensitivity (318 more and 291 less sensitive) between CR and AD cells in a non-androgen environment. 
However, only 108 of these shifts were greater than one standard deviation above the mean. 
In CR cells, HER2 activation of ERK and PI3K was more sensitive, as was AR activation through the MAPK pathway. 
This was expected, as outlaw pathway activity was elevated in castration resistant cells. 
Species in the MAPK pathway were in general more sensitive in CR cells (128 out of 140 significant), with all forms of sPAcP more robust in CR cells.
On the other hand, infrastructure pathways encoding transcription and translation were more sensitive in AD cells.
PSA  and cyclin D1b (mRNA and mRNA complexes) were the only species involved in translation that were more robust in AD cells (14 out of 116).
The transcription factor, E2F was more fragile in AD cells, while the transcription factors ETS and AP1 were more robust. 
ETS and AP1 are activated by phosphorylated ERK, and ETS is also activated by active PKC \cite{Wilkinson2000,Lindemann2003}. 
E2F is deactivated through binding to Rb, which is deactivated by cyclin D1 and CDK phosphorylation \cite{Lapenna2009}. 
The model also included AP1 suppression of AR transcriptional activity (more sensitive in CR) \cite{Sato1997}, as well as inhibition of transcription of the AR gene by E2F (more sensitive in AD) \cite{Davis2006}.
Species in the PI3K pathway that were more fragile in AD cells included Rheb and TOR complexes. 
Interestingly, these species were included as the last step in the PI3K pathway prior to translation, with the phosphorylation of 4E-BP1 by TOR being considered the beginning of translation in this model. 
This again indicates that in the absence of DHT general translation is more fragile in AD cells.

There was a large shift in sensitive species between an androgen and a non-androgen environment in both AD and CR cell lines (Fig. \ref{fg:Sensitivity}C and Fig. \ref{fg:Supp_Sensitivity}). 
Of the 664 statistically significant shifts in AD cells, 288 were more sensitive between androgen versus non-androgen environments. 
However, only 119 shifts were larger than one standard deviation above the mean. 
Unsurprisingly, AR activation through DHT binding, with and without coactivators, in a DHT environment was more sensitive, as was AR inhibition of PAcP transcription (repressed by AR in the model). 
Species further upstream, such as HER2 activation of the MAPK and PI3K/AKT pathways, were also more sensitive in a DHT environment. 
Cell cycle species that were more fragile in the presence of DHT, included complexes involving p21Cip1 and CDC25A.
In a non-androgen environment, basal transcription (68 out of 72) and translation (114 out of 120) were more fragile. 
Other fragile species in the absence of DHT included Rb, E2F, Sam68, cyclin D1a complexes, phosphatases in the MAPK pathway, Rheb complexes, and TOR complexes.

We also considered the sensitivity of CR cells following the application of the AR inhibitor enzalutamide in the presence of DHT (Fig. \ref{fg:Sensitivity}D). 
In an androgen environment with enzalutamide, species which were more fragile included cytosolic AR, cPAcP, and p21Cip1. 
As we would expect, AR species found in the nucleus and/or bound to coactivators, were more robust in the absence of enzalutamide. 
The top two percent of sensitive species with and without enzalutamide were conserved. 
In a CR cell, enzalutamide had no effect on the sensitivity of PI3K/AKT species as well as many MAPK species (ERK, Raf, and MEK). 
Next, we looked at the effect of enzalutamide on a CR cell in both a non-androgen and DHT environment (Fig. \ref{fg:Supp_Sensitivity}). 
More sensitive species in a non-androgen environment included dimerized HER2, ERK, and PAcP. 
Species which were more robust in the non-androgen environment included, AR activated by DHT, AKT, p70, and AR bound to HSP.
The results of our sensitivity analysis indicate that instead of inhibiting solely the AR pathway (enzalutamide), a combination therapy targeting the PI3K or MAPK pathways in addition to AR may be more effective.

\subsection*{Robustness analysis identified key regulators of prostate cancer.} 
Robustness analysis was conducted for 80 proteins to quantify the effects of amplifying or knocking down key model components in both AD and CR cells. 
Gene expression parameters were altered by a factor 10, 0.5, and 0 for knock-in, knock-down, or knock-out perturbations, respectively.
We calculated the effect of these perturbations on different protein markers, such as PSA, AR, and cyclin D. 
The results of the model suggest that an inhibition of either the PI3K pathway or the MAPK pathway in combination with an AR inhibitor as a possible therapy for CRPC. 
Sensitivity analysis revealed no change in the top sensitive species in the presence or absence of the AR inhibitor, enzalutamide. PI3K/AKT and MAPK species continued to fall in the top two percent of sensitive species. 
A study by Carver \textit{et al.} looked at dual inhibition of AR and PI3K signaling in LNCaP cells and in a Pten-deficient murine prostate cancer model \cite{Parker2013}. 
Using both the PI3K inhibitor, BEZ235, and the AR inhibitor, MDV3100 (enzalutamide), the group saw a drastic decrease in the total number of cells. 
Each inhibitor on it’s own had a much smaller effect on total cell number. 
They saw an increase in the cell death marker, c-PARP, in the dual inhibition case. 
The group hypothesized that AKT inhibition leads to increased AR signaling activity through increased protein concentrations of HER3. 
On the other hand, AR inhibition leads to increased AKT activity due to the down regulation of PHLPP, a protein phosphatase that regulates AKT. 
For simplicity, the HER3 pathway and also cell death were not included in the model. 
Dual knock-out studies of PI3K and AR in our model show no additive effect on any cell cycle proteins through the dual knock-out. 
The pathways in our model appear to be uncoupled and therefore no synergy is shown in the dual knock-out case. 
This could indicate that the combined decrease in cell population is entirely due to cell death. 
The Carver \textit{et al.} study did not consider cell cycle proteins or cell growth. 
Our model does show a decrease in cell cycle proteins in the PI3K knock-out as well as in the dual knock-out case. 
This result seems to be consistent with the decreased cell growth in the PI3K knock-out case which is not dependent on cell death, as c-PARP levels are low. 
The decrease in cell cycle proteins is due to a decrease in general translation, including free eIF4E levels and activated 40S ribosome subunit. 
The decrease in p70 (S6) activation due to inhibition of PI3K is shown in both the model and in the Carver \textit{et al.} study (Supplementary Figure), indicating this result is due to the PI3K pathway. 
It would be interesting to repeat the experiment looking at cell cycle proteins and also performing the experiment in CR LNCaP cells, instead of AD cells.
